Future challenges for Research Software Engineering

### 2025-09-25 future_challenges_for_RSEng.tex
Update future_challenges_for_RSEng.tex
Co-authored-by: Gerasimos Chourdakis <gerasimos.chourdakis@ipvs.uni-stuttgart.de>
fixing pronouns and tenses (#69)
* feat(#44): fix pronouns

* feat(#44): fix tense

* swap words

* split long sentence

* replace word

* drop paragraph as it is already a few lines above

* reduce the number of certains

* Update future_challenges_for_RSEng.tex

Co-authored-by: Gerasimos Chourdakis <gerasimos.chourdakis@ipvs.uni-stuttgart.de>

Research Software Engineering is still a developing field,
with increasing integration into various aspects of research.
In Germany, a formal call for creating a sustainable environment for research software has only been raised by the RSE community a couple of years ago~\cite{Anzt2021},
while the foundational competencies of an RSE have just recently been defined in detail~\cite{Goth2024}.
Central RSE units are in the process of being imagined~\cite{Kempf2025-draft},
while an RSE Master's is in the early stages of design~\cite{Dehne2025-draft}.

In this opinion paper, we try to imagine the state of our field
and community in the mid-term future, focusing on future challenges.
While future projections are standard practice in software engineering (SE)~\cite{Katz2023,Khan2021,Hu2023,Bosch2016a,Boehm2011},
so far future projections are far less common for the RSEng domain.
Here, we present the results of a workshop discussion (notes content is archived here \cite{goth2025_WSPad}) at the
deRSE25 conference in Karlsruhe\footnote{deRSE25, session ``The End of {RSEng}?
Challenges and Risks for {RSEng}'': \url{https://events.hifis.net/event/1741/contributions/14026/}},
without claiming to represent the entire deRSE community.
The material presented here is based on the output of the workshop participants, but extended with the authors' ideas.
Since deRSE25 was co-located with SE25, the annual conference for the Software Engineering community in Germany, the discussion was also open to software engineers,
but no list or profile of the participants is available.
The workshop hosts felt, that the majority of participants were from the RSE community.
Nevertheless, by disseminating the discussed ideas,
we aim to engage the community in creating a future worth living
and to advance the conversation about how RSEng should, or could, develop.

@@

\begin{story}{The greater world of Kim and Kay}{ethics}
10 years have passed since Kim and Kay started on their journey of being RSEs. Back then,
at the universities of Eden and Utopia, life was more orderly and harmonious.
Kim is currently designing the software service stack at the university.
She increasingly struggles to find compelling Open Source alternatives that enables them to keep complete control of their data
rather than relying on software and tools that are backed by large corporate entities.
So far, she still has the support of the university board, since it values strategic autonomy.
Kay had left academia to work in an industrial R\&D department because of a more reliable employment.
There she gained valuable experience working in an industrial setting.
However, she was very happy when an opportunity arose to work in a central
RSE department with an open-ended contract. Apart from her main responsibility
of developing and maintaining software for an affordable water monitoring system
used in the global south, she now also teaches undergraduates digital literacy.
\end{story}

>>>>

Research Software Engineering is still a developing field,
with increasing integration into various aspects of research.
In Germany, a formal call for creating a sustainable environment for research software
was only published a couple of years ago~\cite{Anzt2021},
while the foundational competencies of an RSE have only recently been defined in detail~\cite{Goth2024}.
Central RSE units are in the process of being imagined~\cite{Kempf2025-draft},
while an RSE Master's is in the early stages of design~\cite{Dehne2025-draft}.

In this opinion paper, we try to imagine the state of our field
and community in the mid-term future, focusing on future challenges.
While future projections are standard practice in software engineering (SE)~\cite{Katz2023,Khan2021,Hu2023,Bosch2016a,Boehm2011},
so far future projections are far less common for the RSEng domain.
Here, we present the results of a workshop discussion (notes content is archived here \cite{goth2025_WSPad}) at the
deRSE25 conference in Karlsruhe\footnote{deRSE25, session ``The End of {RSEng}?
Challenges and Risks for {RSEng}'': \url{https://events.hifis.net/event/1741/contributions/14026/}},
without claiming to represent the entire deRSE community.
The material presented here is based on the output of the workshop participants, but extended with the authors' ideas.
Since deRSE25 was co-located with SE25, the annual conference for the Software Engineering community in Germany, the discussion was also open to software engineers,
but no list or profile of the participants is available.
The workshop hosts felt, that the majority of participants were from the German RSE community.
Nevertheless, by disseminating the discussed ideas,
we aim to engage the community in creating a future worth living
and to advance the conversation about how RSEng should, or could, develop.
Since this paper focuses on the outputs of the deRSE25 conference, we present our thoughts and perspectives in a largely German-focused context.
At the same time, we recognise, the wider international growth of RSEng as a very important development
for people undertaking software development within the research community.
Much of our work and our discussions are also building on wider perspectives for the future of RSEng that are currently under
investigation among RSEng communities in other countries.
While taking a German-focused view in this paper, we therefore aim to highlight resources and materials from
the wider international community where we feel this is relevant.

@@

\begin{story}{The greater world of Kim and Kay}{ethics}
10 years have passed since Kim and Kay started on their journey of being RSEs. Back then,
at the universities of Eden and Utopia, life was more orderly and harmonious.
Kim is currently designing the software service stack at the university.
She increasingly struggles to find compelling Open Source alternatives that enables them to keep complete control of their data
rather than relying on software and tools that are backed by large corporate entities.
So far, she still has the support of the university board, since it values strategic autonomy.
Kay had left academia to work in an industrial R\&D department because of a more reliable employment.
There, she gained valuable experience working in an industrial setting.
However, she was very happy when an opportunity arose to work in a central
RSE department with an open-ended contract. Apart from her main responsibility
of developing and maintaining software for an affordable water monitoring system
used in the global south, she now also teaches undergraduates digital literacy.
\end{story}
